Seg_sites_rarefaction

### 2024-02-15 scripts/seg_sites_rarefaction.Rmd
work on figure revisions

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE); library(snpR); library(ggplot2); library(data.table);
```

# Define parameters
```{r}
# define parameters
seed <- 231233
threads <- 25
allele_frequencies <- seq(.001, .1, by = .001) # allele frequencies
n1 <- 5000 # p1 gene copy number
n2 <- 500 # p2 gene copy number
k <- seq(10, 100, by = 10) # numbers to rarefact to
mcmcs <- seq(1000, 10000, by = 1000) # number of mcmc draws
max_missing <- 30 # maximum percentage of missing data in a locus, from a uniform dist

@@

# import and condense simulation results
```{r}
res_files <- list.files("../data/seg_sites/", "\\.RDS", full.names = TRUE)
res_files <- lapply(res_files, readRDS)
ks <- unlist(purrr::map(res_files, "k"))
n_iters <- unlist(purrr::map(res_files, "iters"))

@@

```{r Figure1}
prep_f1 <- function(pseg1, pseg2, mpseg1, mpseg2){
  #========plot the emperical CIs around every loci vs the predicted value==============
  # plotting the emperical CIs vs the estiamted mean
  pseg <- rbind(data.frame(pop = paste0("N = ", n1/2), p = pseg1, snp = 1:nal, af = allele_frequencies),
                data.frame(pop = paste0("N = ", n2/2), p = pseg2, snp = 1:nal, af = allele_frequencies))

@@

K10_data <- which(ks == 10 & n_iters == 10000)
F1data_K10 <- prep_f1(pseg1 = pseg1$K_10, 
                      pseg2 = pseg2$K_10, 
                      mpseg1 = res_files[[K10_data]]$mpseg1,
                      mpseg2 = res_files[[K10_data]]$mpseg2)

K100_data <- which(ks == 100 & n_iters == 10000)
F1data_K100 <- prep_f1(pseg1 = pseg1$K_100, 
                       pseg2 = pseg2$K_100, 
                       mpseg1 = res_files[[K100_data]]$mpseg1,
                       mpseg2 = res_files[[K100_data]]$mpseg2)

@@

```{r}
prep_f2 <- function(seg1, seg2, vseg1, vseg2, F1data){
  seg <- F1data

  nap_pi <- function(alpha, vseg, n) qnorm(1 - (alpha/2))*sqrt(vseg)*sqrt(1 + (1/n))
  nap_ci <- function(alpha, vseg, n) qnorm(1 - (alpha/2))*sqrt(vseg)/sqrt(n)

@@

K10_data <- which(ks == 10 & n_iters == 10000)
F2data_K10 <- prep_f2(seg1 = res_files[[K10_data]]$seg1, 
                      seg2 = res_files[[K10_data]]$seg2, 
                      vseg1 = vseg1$K_10,
                      vseg2 = vseg2$K_10,
                      F1data = F1data_K10)

K100_data <- which(ks == 100 & n_iters == 10000)
F2data_K100 <- prep_f2(seg1 = res_files[[K100_data]]$seg1, 
                       seg2 = res_files[[K100_data]]$seg2, 
                       vseg1 = vseg1$K_100,
                       vseg2 = vseg2$K_100,
                       F1data = F1data_K100)

@@

f2 <- ggplot(F2data_mcmcs, aes(x = nseg)) +
  geom_density() +
  geom_vline(data = F2data_sums, aes(xintercept = nseg), color = colors[1]) +
  geom_vline(data = F2data_pi, aes(xintercept = lower), linetype = "dashed", color = colors[2]) +
  geom_vline(data = F2data_pi, aes(xintercept = upper), linetype = "dashed", color = colors[2]) +
  facet_grid(K~pop, labeller = label_parsed) +
  theme_bw() +
  scale_y_continuous(expand = c(0,0)) +
  theme(axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        strip.text = element_text(size = t.sizes[1]),
        strip.background = element_blank()) +
  xlab(bquote(italic(E*"("*N[S]*")"))) +
  ylab("Density")

Figure2 <- ggplot(sum_tab, aes(x = pop, y = nseg, color = in_ci)) + 
  geom_jitter(data = all_mcmc_sums, aes(x = pop, y = nseg, color = in_pi), height = 0, alpha = .2) +
  geom_point(size = 4, shape = 17) +
  geom_errorbar(data = sum_tab, aes(ymax = upper, ymin = lower, x = pop), 
                inherit.aes = FALSE, width = .25, color = ci_col, linewidth = 1) +
  geom_errorbar(data = pi_tab, aes(ymax = upper, ymin = lower, x = pop), 
                inherit.aes = FALSE, width = .5, linetype = "dashed", color = ci_col, linewidth = 1) +
  theme_bw() +
  theme(strip.background = element_blank()) +
  scale_color_manual(values = colors) +
  guides(color = guide_legend(title = "In CI/PI")) +
  ylab(bquote(italic(E*"("*N[S]*")"))) +
  xlab("Population")

@@

tapply(all_mcmc_sums$in_pi, all_mcmc_sums$pop, mean, na.rm = TRUE)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE); library(snpR); library(ggplot2); library(data.table);
library(cowplot)
```

# Define parameters
```{r}
# define parameters
seed <- 231266
threads <- 25
allele_frequencies <- seq(.001, .1, by = .001) # allele frequencies
n1 <- 5000 # p1 gene copy number
n2 <- 500 # p2 gene copy number
k <- seq(10, 100, by = 10) # numbers to rarefact to
mcmcs <- seq(1000, 10000, by = 1000) # number of mcmc draws
max_missing <- 30 # maximum percentage of missing data in a locus, from a uniform dist

@@

# import and condense simulation results
```{r}
res_files <- list.files("../data/", "\\.RDS", full.names = TRUE)
res_files <- lapply(res_files, readRDS)
ks <- unlist(purrr::map(res_files, "k"))
n_iters <- unlist(purrr::map(res_files, "iters"))

@@

```{r Figure1}
prep_f1 <- function(pseg1, pseg2, mpseg1, mpseg2, n1, n2){
  #========plot the emperical CIs around every loci vs the predicted value==============
  # plotting the emperical CIs vs the estiamted mean
  pseg <- rbind(data.frame(pop = paste0("N = ", n1/2), p = pseg1, snp = 1:nal, af = allele_frequencies),
                data.frame(pop = paste0("N = ", n2/2), p = pseg2, snp = 1:nal, af = allele_frequencies))

@@

K10_data <- which(ks == 10 & n_iters == 10000)
F1data_K10 <- prep_f1(pseg1 = pseg1$K_10, 
                      pseg2 = pseg2$K_10, 
                      mpseg1 = res_files[[K10_data]]$mpseg1,
                      mpseg2 = res_files[[K10_data]]$mpseg2,
                      n1 = n1,
                      n2 = n2)

K100_data <- which(ks == 100 & n_iters == 10000)
F1data_K100 <- prep_f1(pseg1 = pseg1$K_100, 
                       pseg2 = pseg2$K_100, 
                       mpseg1 = res_files[[K100_data]]$mpseg1,
                       mpseg2 = res_files[[K100_data]]$mpseg2,
                       n1 = n1,
                       n2 = n2)

@@

```{r}
prep_f2 <- function(seg1, seg2, vseg1, vseg2, F1data, n1, n2){
  seg <- F1data

  # normal approximation CI and PI equation given var and N
  nap_pi <- function(alpha, vseg, n) qnorm(1 - (alpha/2))*sqrt(vseg)*sqrt(1 + (1/n))
  nap_ci <- function(alpha, vseg, n) qnorm(1 - (alpha/2))*sqrt(vseg)/sqrt(n)

@@

K10_data <- which(ks == 10 & n_iters == 10000)
F2data_K10 <- prep_f2(seg1 = res_files[[K10_data]]$seg1, 
                      seg2 = res_files[[K10_data]]$seg2, 
                      vseg1 = vseg1$K_10,
                      vseg2 = vseg2$K_10,
                      F1data = F1data_K10,
                      n1 = n1,
                      n2 = n2)

K100_data <- which(ks == 100 & n_iters == 10000)
F2data_K100 <- prep_f2(seg1 = res_files[[K100_data]]$seg1, 
                       seg2 = res_files[[K100_data]]$seg2, 
                       vseg1 = vseg1$K_100,
                       vseg2 = vseg2$K_100,
                       F1data = F1data_K100,
                       n1 = n1,
                       n2 = n2)

@@

f2 <- ggplot(F2data_mcmcs, aes(x = nseg)) +
  geom_density() +
  geom_vline(data = F2data_sums, aes(xintercept = nseg), color = colors[1]) +
  geom_vline(data = F2data_pi, aes(xintercept = lower), linetype = "dashed", color = colors[2]) +
  geom_vline(data = F2data_pi, aes(xintercept = upper), linetype = "dashed", color = colors[2]) +
  facet_grid(K~pop, labeller = label_parsed) +
  theme_bw() +
  scale_y_continuous(expand = c(0,0)) +
  theme(axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        strip.text = element_text(size = t.sizes[1]),
        strip.background = element_blank()) +
  xlab(bquote(italic(N[S]))) +
  ylab("Density")

@@

# Figure 3: var vs difference between estimate and mean of simulations for mcmc and k

```{r Figure3}
# for all of the mpseg files, figure out if the percentage of points in the CI changes
## do for both each SNP and for Ns
ps_list <- ns_list <- vector("list", length(res_files))
for(i in 1:length(res_files)){
  tk <- res_files[[i]]$k
  titers <- res_files[[i]]$iters

  ps_prep <- prep_f1(pseg1 = pseg1[[paste0("K_", tk)]],
                     pseg2 = pseg2[[paste0("K_", tk)]], 
                     mpseg1 = res_files[[i]]$mpseg1,
                     mpseg2 = res_files[[i]]$mpseg2,
                     n1 = n1,
                     n2 = n2)

  ns_prep <- prep_f2(seg1 = res_files[[i]]$seg1, 
                     seg2 = res_files[[i]]$seg2, 
                     vseg1 = vseg1[[paste0("K_", tk)]],
                     vseg2 = vseg2[[paste0("K_", tk)]],
                     F1data = ps_prep,
                     n1 = n1,
                     n2 = n2)

  ps_prep$K <- tk
  ps_prep$iters <- titers
  ps_prep$sim_num <- i

  ns_prep$all_mcmc_sums$K <- tk
  ns_prep$all_mcmc_sums$iters <- titers
  ns_prep$all_mcmc_sums$i <- i
  ns_prep$all_mcmc_sums$ENs <- ns_prep$sum_tab$nseg[match(ns_prep$all_mcmc_sums$pop,
                                                          ns_prep$sum_tab$pop)]

  ps_list[[i]] <- ps_prep
  ns_list[[i]] <- ns_prep$all_mcmc_sums
}

ps_all <- rbindlist(ps_list)

ps_all_m <- ps_all[source == "Mathematical",]
ps_all_s <- ps_all[source == "Simulation",]
ps_all_merge <- merge(ps_all_m[,-c("source")], 
                      ps_all_s[,-c("source")], 
                      by = c("af", "pop", "snp", "sim_num", "lower", "upper", "K", "iters"), 
                      suffixes = c(".mathematical", ".simulated"))
ps_all_merge[,diff := p.mathematical - p.simulated]

# Figure S1: gamma vs error rate
ps_all_aggregate_for_K <- ps_all_merge[, lapply(.SD, mean), .SDcols = "in_ci.mathematical", by = .(K, iters, pop)]
FS1 <- ggplot(ps_all_aggregate_for_K, aes(x = K, y = in_ci.mathematical*100)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(iters~pop) +
  theme_bw() +
  # khroma::scale_color_batlow(discrete = TRUE, reverse = TRUE) +
  # khroma::scale_fill_batlow(discrete = TRUE, reverse = TRUE) +
  theme(axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        strip.text = element_text(size = t.sizes[1]),
        strip.background = element_blank(),
        legend.title = element_text(size = t.sizes[1]),
        legend.text = element_text(size = t.sizes[2])) +
  xlab(parse(text = "gamma")) +
  ylab("% estimates in CI")

# Figure S2: maf vs error rate
ps_all_aggregate_for_af <- ps_all_merge[, lapply(.SD, mean), .SDcols = "in_ci.mathematical", by = .(af, iters, pop)]
FS2 <- ggplot(ps_all_aggregate_for_af, aes(x = af, y = in_ci.mathematical*100)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(iters~pop) +
  theme_bw() +
  # khroma::scale_color_batlow(discrete = TRUE, reverse = TRUE) +
  # khroma::scale_fill_batlow(discrete = TRUE, reverse = TRUE) +
  theme(axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        strip.text = element_text(size = t.sizes[1]),
        strip.background = element_blank(),
        legend.title = element_text(size = t.sizes[1]),
        legend.text = element_text(size = t.sizes[2])) +
  xlab("Minor Allele Frequency") +
  ylab("% estimates in CI")

f3a <- ggplot(ps_all_merge, aes(x = K, y = diff, color = in_ci.mathematical)) +
  geom_point() +
  theme_bw() +
  facet_grid(~pop) +
  khroma::scale_color_highcontrast() +
  scale_x_continuous(breaks = k) +
  theme(axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        strip.text = element_text(size = t.sizes[1]),
        strip.background = element_blank(),
        legend.title = element_text(size = t.sizes[1]),
        legend.text = element_text(size = t.sizes[2])) +
  xlab(parse(text = "gamma")) +
  ylab(bquote(italic(P*"("*S[jq]*")"[E]) - italic(P*"("*S[jq]*")"[O]))) +
  guides(color = guide_legend(title = "In 95% CI"))

f3b <- ggplot(ps_all_merge, aes(x = af, y = diff, color = in_ci.mathematical)) +
  geom_point() +
  theme_bw() +
  facet_grid(~pop) +
  khroma::scale_color_highcontrast() +
  theme(axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        strip.text = element_text(size = t.sizes[1]),
        axis.text.x = element_text(size = t.sizes[2], angle = 90, vjust = .5),
        strip.background = element_blank(),
        legend.title = element_text(size = t.sizes[1]),
        legend.text = element_text(size = t.sizes[2])) +
  xlab("Minor Allele Frequency") +
  ylab(bquote(italic(P*"("*S[jq]*")"[E]) - italic(P*"("*S[jq]*")"[O])))

f3c <- ggplot(ps_all_merge, aes(x = iters, y = diff, color = in_ci.mathematical)) +
  geom_point() +
  theme_bw() +
  facet_grid(~pop) +
  khroma::scale_color_highcontrast() +
  scale_x_continuous(breaks = mcmcs) +
  theme(axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        strip.text = element_text(size = t.sizes[1]),
        axis.text.x = element_text(size = t.sizes[2], 
                                   angle = 90, vjust = .5),
        strip.background = element_blank(),
        legend.title = element_text(size = t.sizes[1]),
        legend.text = element_text(size = t.sizes[2])) +
  xlab("Number of Simulated Draws") +
  ylab(bquote(italic(P*"("*S[jq]*")"[E]) - italic(P*"("*S[jq]*")"[O])))

f3l <- ggpubr::get_legend(f3a)
f3 <- plot_grid(f3a + guides(color = "none"), 
          f3b + guides(color = "none"), 
          f3c + guides(color = "none"),
          ncol = 1, 
          align = "bl", 
          axis = "bl")

f3 <- plot_grid(f3, f3l, nrow = 1, rel_widths = c(1, .2))

save_plot("Figure3.pdf", f3, base_height = 11, base_width = 15)
```
# Figure 4 Nseg vs parameters

```{r Figure4}
ns_all <- rbindlist(ns_list)

ns_all$diff <- ns_all$ENs - ns_all$nseg

f4a <- ggplot(ns_all[iters == 10000,], aes(x = K, y = nseg)) +
  geom_errorbar(data = unique(ns_all[iters == 10000, c("pop", "K", "iters", "i", "lower", "upper")]),
            aes(x = K, ymin = lower, ymax = upper), color = "#DDAA33", size = 1, 
            position = position_nudge(x = 3.5),
            inherit.aes = FALSE, width = 2) +
  stat_ydensity(aes(group = K), draw_quantiles = c(.025, .975), scale = "width", bw = .4,
                fill = NA, width = 6) +
  facet_wrap(~pop) +
  theme_bw() + 
  scale_x_continuous(breaks = k) +
  theme(axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        legend.title = element_text(size = t.sizes[1]),
        strip.text = element_text(size = t.sizes[1]),
        strip.background = element_blank(),
        legend.text = element_text(size = t.sizes[2])) +
  ylab(bquote(italic(E*"("*N[S]*")") - italic(N[S,observed]))) +
  xlab(bquote(gamma))

f4b <- ggplot(ns_all[K == 50,], aes(x = iters, y = nseg)) +
  geom_errorbar(data = unique(ns_all[K == 50,c("pop", "K", "iters", "i", "lower", "upper")]),
            aes(x = iters, ymin = lower, ymax = upper), color = "#DDAA33", size = 1, 
            position = position_nudge(x = 350),
            inherit.aes = FALSE, width = 200) +
  stat_ydensity(aes(group = iters), draw_quantiles = c(.025, .975), scale = "width", bw = .4,
                fill = NA, width = 600) +
  facet_wrap(~pop) +
  theme_bw() + 
  scale_x_continuous(breaks = mcmcs) +
  theme(axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        legend.title = element_text(size = t.sizes[1]),
        strip.text = element_text(size = t.sizes[1]),
        strip.background = element_blank(),
        legend.text = element_text(size = t.sizes[2])) +
  ylab(bquote(italic(E*"("*N[S]*")") - italic(N[S,observed]))) +
  xlab("Number of Simulated Draws")
